Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-05-19 Ben_Manuscripts/hdphmm/Draft.tex
moving figures to SI

  Appropriate time series modeling of complex diffusion in soft matter systems on the
  microsecond time scale can provide a path towards inferring transport mechanisms and
  predicting bulk properties characteristic of much longer time scales. In this work 
  we apply the sticky hierarchical Dirichlet process hidden Markov model (HPD-AR-HMM) 
  %MRS: might be confusing to have the AR in there when the words ``Auto-regressive'' 
  %don't appear; then they would try to figure out where the AR is. 
  to solute center-of-mass trajectories generated from long molecular dynamics 
  simulations in a cross-linked inverted hexagonal phase lyotropic liquid crystal
  (LLC) membrane in order to automatically detect a variety of solute dynamical states
  which can be further explained in terms of solute-membrane interactions.

@@

  %MRS: we have to decide if ``effective'' is the right word. Alternative is ``promising''.
  Overall, this work provides an effective way to simultaneously identify transport 
  mechanisms in nanoporous materials and project complex diffusive behavior on
  long time scales.  
  %MRS: simplify by removing nominalizations.
  %This enhancement to our understanding 
  Our enhanced understanding
  of the diverse range 
  of solute behavior allows us to hypothesize design changes 
  to LLC monomers aimed towards controlling the rates of solute passage, thus improving 
  the selective performance of LLC membranes.

@@

  In this work, we are particularly interested in the design of lyotropic liquid 
  crystal (LLC) monomers, a class of amphiphilic molecules whose ordered self-assembled
  phases can be cross-linked into mechanically strong membranes capable of highly 
  selective separations. Inverted hexagonal, H\textsubscript{II}, phase LLC membranes
  are characterized	by hexagonally packed, uniform-sized and straight pores, an ideal
  geometry for high throughput transport. Their pores are lined with the LLC monomer's
  functional groups which can potentially be designed to interact with solutes in a 
  chemically-specific manner. It would be extremely useful to experimentalists if we 
  could connect LLC monomer design with measurable quantities. This goal is complicated
  by the non-Brownian hopping and trapping behavior induced by the membrane's diverse
  topology and inhomogeneous structure result in subdiffusive behaviour out to multiple microseconds of simulation.~\cite{coscia_understanding_2019,coscia_chemically_2019}
  %MRS: to sentence above added to come back to main theme

  We have already advanced our understanding of LLC monomer design in our previous work
  by building stochastic time series models based on the chemical intuition gained
  from qualitative MD studies of solute transport in an H\textsubscript{II} phase LLC
  membrane.~\cite{coscia_chemically_2019,coscia_capturing_2020} We identified three 
  classes of trapping behavior which result in 
  %MRS: tweaked
  subdiffusion out to the microsecond time scale. 
  In our 
  first approach to stochastically model this behavior, we treated the system's dynamics
  as a sequence of anti-correlated hops between power law distributed periods of 
  entrapment, a framework called subordinated fractional Brownian or L\'evy
  motion.~\cite{thiel_weak_2014,teuerle_modeling_2013} In a second approach, we treated 
  solute motion as a Markov state model~\cite{pande_everything_2010} with state-dependent
  dynamics, where we parameterized the state transition probabilities between each of
  eight discrete states, defined by the observed trapping mechanisms, as well as the 
  solute dynamics within each of these states. We showed how one could use realizations
  of any stochastic model like these in order to predict macroscopic flux and selectivity.

  Although powerful and insightful, our previous work required considerable human effort
  in order to analyze the MD trajectories and to hypothesize stochastic models which
  matched the behavior of solutes. It would be significantly more useful and broadly 
  applicable if we could design an approach which 
  %MRS: tweaked to sound more machine-learning-y
  %automatically 
  distinguishes and 
  parameterizes varied solute dynamics 
  in a data-driven way
  and leaves a facile way to generate an ensemble
  of stochastic trajectory realizations that can be used to predict macroscopic transport
  properties.

@@

  In their most basic form, hidden Markov models (HMMs) are used to generate
  a series of latent, or hidden, states dependent only on the previous state.~\cite{rabiner_tutorial_1989}
  Propagating an HMM generates sequential but independent observations from 
  the emission distributions associated with each hidden state. Transitions 
  between states are mathematically defined by a matrix of transition probabilities.
  Given a series of observations and a known number of hidden states, the 
  Baum-Welch algorithm is the most commonly used way to determine the hidden
  state sequence and the underlying parameters of a standard HMM.~\cite{baum_maximization_1970}
  It also possible to achieve the same results using a Bayesian 
  framework.~\cite{scott_bayesian_2002,jasra_markov_2005} 
  %MRS: not entirely clear what purpose of above sentence its - it sort of implied you will use Bayesian framework?
  One can place priors
  on the transition probabilities and parameters of the hidden states' emission
  distributions in order to iteratively infer their values and to estimate the hidden state
  sequence. 
  HMMs have been used in a broad range of applications including speech
  recognition, weather forecasting and bioinformatics.~\cite{juang_hidden_1984,hughes_non-homogeneous_1999,yoon_hidden_2009}.
  In molecular simulations, HMMs have been
  used to characterize molecular kinetics in order to determine the equilibrium 
  distribution of various observables.~\cite{noe_probability_2008}
  %MRS: probably a few more citations that should be added above, since there has been a lot.
  By placing a hierarchical Dirichlet process (HDP) prior on the transition 
  probabilities, it is possible to extend HMMs to an unknown number of hidden
  states, thus allowing the model's complexity to be driven by the complexity
  of the data.~\cite{teh_hierarchical_2006} We will refer to this as the HDP-HMM,
  although in literature it is often called the infinite hidden Markov 
  model.~\cite{beal_infinite_2002} The HDP-HMM elegantly overcomes the main
  shortcoming of standard HMMs, the need to choose the number of states 
  \textit{a priori}, making its use more advantageous in a diversity of 
  applications including economic data and single particle trajectory 
  analysis.~\cite{shi_identifying_2016,hines_analyzing_2015}

@@

  %MRS: define what ``this'' is - important claim, make it clear what it is you are claiming.
  To our knowledge, this is a novel application and analysis of the sticky 
  HDP-AR-HMM. Typical applications center around determining the number of
  states, the state sequence and occasional analysis of the emission 
  distributions. The addition of temporal correlation to the inference 
  procedure has primarily been used to improve state sequence 
  identification.~\cite{calderon_inferring_2015,hamada_modeling_2016}
  Not only do we use the HDP-AR-HMM to segment MD time series into distinct
  dynamical modes, but we cluster and analyze the temporally-correlated 
  parameters of the model in order to learn physical mechanisms and describe
  solute kinetics. Further, we use the model parameters to generate new 
  time series trajectories which enable projections of MD simulations on 
  much longer time scales.

@@

  We ran all MD simulations and energy minimizations using GROMACS 2018.~\cite{bekker_gromacs:_1993,berendsen_gromacs:_1995,van_der_spoel_gromacs:_2005,hess_gromacs_2008}  
  Our Python implementation of the HDP-AR-HMM algorithm is available online at \\
  %MRS: possibly should give the permanent reference to a shirtsgroup fork?  We can discuss which is likely to be more permenant
  \texttt{https://github.com/bencoscia/hdphmm}. All other post-simulation 
  trajectory analysis tools are available online at
  \texttt{https://github.com/shirtsgroup/LLC\_Membranes}.

@@

  %MRS: tweaked
  %We let each solute undergo 
  Each solute system was simulated for 5 $\mu$s of molecular dynamics  simulation. We used a time step of 2 fs
  at a pressure of 1 bar and temperature of 300K controlled by the Parrinello-Rahman 
  %MRS: just wondering, was this semiisotropic, or isotropic?  If isotropic, do we know how different the diagonals of the 
  pressure tensor were (since the system is not totally isotropic)
  barostat~\cite{parrinello_polymorphic_1981} and velocity rescale 
  thermostat~\cite{bussi_canonical_2007} respectively. We recorded configurations every 0.5 ns.

@@

  % BJC4: Chris' general comment for this section: 'the problem with this section is 
  % that it doesn't help the expert or novice figure out what is being done (either 
  % expand more or relegate details to references and talk bigger picture)'

  % BJC4: I think it makes sense to stick to bigger picture since there are much
  % better descriptions in literature. I haven't made any modifications to what's already been done

  % BJC4: replaced with what follows. Lots of the big picture is in the intro

%  \subsubsection*{Extending Hidden Markov Modeling to an Unknown Number of States}
%  
%  Hidden Markov models (HMMs) are a useful and widely used technique for modeling
%  sequences of observations where the probability of the next observation in a 
%  sequence depends, at least in part, on a previous unobserved, latent or hidden,
%  state.~\cite{beal_infinite_2002} In the context of our simulations, the observations
%  correspond to the center of mass coordinates of the solutes versus time, and the
%  states correspond to segments of the trajectory with similar dynamical behavior 
%  which gives rise to those types of observations. The probability of transitioning 
%  to a state based on the current state is defined in terms of an $n\times n$ 
%  transition probability matrix, $T$, where $n$ is the number of states. Unfortunately,
%  standard HMMs require $n$ to be known \textit{a priori}. One can partially overcome
%  this by testing a range of numbers of hidden states and determining which is the
%  best representation of the data.~\cite{pohle_selecting_2017}
%  
%  The infinite-state HMM overcomes this drawback by placing a hierarchical
%  Dirichlet process (HDP) prior on the transition probabilities.~\cite{fox_bayesian_2010} Using some 
%  base probability distribution, $H$, a Dirichlet process (DP) generates discrete 
%  distributions, $G_0$, over a countably infinite number of probability measures:
%  \begin{equation}
%      G_0 = \sum_{k=1}^{\infty} \beta_k \delta_{\theta_k} ~~ \theta_k \sim H, \beta \sim GEM(\gamma)
%  \end{equation}
%  where the $\theta_k$ are values drawn from the base distribution and the weights
%  $\beta_k$ come from a stick-breaking process parameterized by the concentration 
%  parameter $\gamma$ (equivalently referred to as GEM($\gamma$)).~\cite{halmos_random_1944} In shorthand, this
%  specification can be expressed as $G_0 \sim DP(\gamma, H)$. The concentration 
%  parameter, $\gamma$, expresses one's confidence in the base distribution $H$.
%  We use a uniform base distribution for $H$. When $\gamma\to 0$, the first 
%  weight of $G_0$, $\beta_1$, approaches unity and for $\gamma\to\infty$, the weights
%  become uniform and $G_0$ closely resembles $H$. Each row, $G_j$, of the transition 
%  matrix is produced by drawing from a DP specified using the $\beta$ vector as a 
%  discrete base distribution and a separate concentration parameter, $\alpha$.
%  \begin{equation}
%      G_j = \sum_{k=1}^{\infty} \pi_{jk} \delta_{\theta_k} ~~ \pi_j \sim DP(\alpha, \beta)
%  \end{equation}
%  This hierarchical specification ensures that the transition probabilities in 
%  each row share the same support points \{$\theta_1$, ..., $\theta_k$\}.
%  Once the model has converged only a finite number of states will have significant
%  sampling.

  We use the sticky HDP-AR-HMM in order to assign the time-correlated positional 
  fluctuations of solute center-of-mass MD trajectories to discrete dynamical modes while 
  simultaneously inferring the VAR parameters of those modes. We use the sticky model
  in order to emphasize self-transitions since we expect solutes to stay trapped in
  states for appreciable amounts of time. We analyze solute motion in terms of their
  cylindrical coordinates, $(r, z)$, with the $z$ axis oriented along the membrane 
  pores and the radial dimension, $r$, measured from the closest pore center.

  Although the discussion will be presented in terms of 2D cylindrical coordinates, we
  applied the HDP-AR-HMM algorithm to 3D solute center-of-mass coordinate trajectories 
  with the radial dimension represented by the solute's $x$ and $y$ coordinates 
  transformed relative to the nearest pore center. By working in Cartesian coordinates,
  we avoid mathematical complexity introduced by change of coordinates into cylindrical
  coordinates while estimating the state sequence.

  %MRS: some of this is a repetition of what you said above.
  We implemented the sticky HDP-AR-HMM in Python 
  (\texttt{https://github.com/bencoscia/hdphmm}) which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_nonparametric_2009} Since we do not incorporate
  any novel methodology to the HDP-AR-HMM inference procedure, we refer the 
  interested reader to an excellent and in-depth description of the formulation of
  the HDP-AR-HMM and related models.~\cite{fox_bayesian_2010}

@@

  We clustered based on the eigenvalues of $A$ and $\Sigma$, the distance from the 
  pore center and the self-transition probabilities, $T_{ii}$, of each state. $A$ 
  and $\Sigma$ capture the correlation and size of fluctuations. $T_{ii}$ is directly
  related to the average time spent in each state. The distance from the closest pore 
  center is likely meaningful since the membrane is radially inhomogeneous from each 
  pore, transitioning from hydrophilic pores to hydrophobic tails. However, the 
  channels are isotropic along the pore axis so we do not 
  %MRS: added this?  Grammar didn't make sense without.
  cluster
  in means in the $z$ dimension.
  We normalized each clustering dimension by its variance and shifted the mean to zero.

  To take advantage of the system's cylindrical symmetry, we converted the parameters that
  are directly related to solute position into cylindrical coordinates before clustering. 
  We replaced the $x$ and $y$ eigenvalues of $\Sigma$ and $A$ with $\lambda_x$ + 
  $\lambda_y$ and $\lambda_x^2$ + $\lambda_y^2$ respectively, which are invariant with 
  rotation around the $z$ axis. 
  %MRS: maybe reduntant and not quite right, since it reduces the dimensionality of the paraameter su
  % bspace from 3 to 2, but the overall paameter space is bigger than 3
  %in both cases reducing the dimensionality from 3 to 2.  
  We clustered on squared means in $r$, $\mu_r^2 = \mu_x^2 + \mu_y^2$, rather than
  $\mu_x$ and $\mu_y$ individually, with the square $\mu_r$ converting it into the 
  same units as $A$ and $\Sigma$.

  To cluster on $T_{ii}$, we first cast it in terms of the expected dwell times. The expected
  value of the number of sequential self-transitions is simply:
  % BJC4: I numbered this equation because I reference it later.
  \begin{equation}  
   E[dwell~time] = \frac{1}{1 - T_{ii}}
   \label{eqn:dwell_times}
  \end{equation}  
  This relationship implies that dwell times approach infinity as $T_{ii}$ approaches
  1. Therefore we found the most success with agglomerative clustering by linearizing this
  relationship and clustering on $-\log(1 - T_{ii})$.

  %BJC4: Added
  There are alternative ways to cluster the data dependent on one's goals. Our goal is
  to derive the simplest set of parameter which adequately describe solute motion. 
  However, one may want a more detailed understand of specific dimensions of the 
  clustering dataset. In this case one can cluster on one or more dimensions independently
  and then enumerate the possible clusters. This would ensure very little overlap of
  clusters but may result in a large number of them. Once one has achieved a better 
  understanding of solute-membrane interactions, perhaps by using the HDP-AR-HMM, it may be
  beneficial to incorporate other physically motivated variables into the clustering
  procedure. For example, one can gain a thorough understanding of the dynamics of
  different kinds of trapping behavior related to hydrogen bonding and sodium ion association.
  %MRS: last sentence is a little weird. Audience may be left without knowing how this could be achieved.
  %\subsubsection*{Choosing the Number of Clusters}

@@

%  We aimed to maintain the highest silhouette score, and thus lowest number of clusters,
%  possible while preserving the dynamics of each trajectory.
%  %verifying that visually distinct states stayed separated. 
%  We verified preservation of the dynamics by comparing the MSDs of stochastic trajectory
%  realizations based on clustered HDP-AR-HMM parameter sets to the pooled MSDs of stochastic 
%  realizations of HDP-AR-HMMs fit to individual trajectories (see Section~\ref{method:realizations}).
%%  We chose the lowest number of clusters that reasonably reproduced the pooled MSDs at 
%%  short and long time lags (See Figure~\ref{S-fig:MET_nclusters} of the Supporting 
%%  Information).

  %\subsubsection*{Obtaining Parameters of Clustered States}

  %BJC3: modified 
  \textit{Obtaining Parameters of Clustered States}: We remapped and fixed the
  state sequence based on the cluster assignments and 
  parameterized the clusters using the inference component of the HDP-AR-HMM algorithm.
  First, we modified the ($x$, $y$, $z$) solute trajectories so that they had a 
  mean of zero, leaving only the fluctuations. Next, we took advantage of the
  rotational invariance of the system by rotating the coordinates about the pore 
  axis such that the mean vector of each separate segment was aligned with the 
  $x$ axis. Next, we subtracted the means from each separate segment in the $x$ 
  and $z$ dimensions, with the mean in $y$ being zeroed by the rotation. We used
  the inference component of the HDP-AR-HMM algorithm on this modified trajectory to 
  infer the clustered state parameters by fixing the clustered state sequence.

@@

  \subsection{Predicting Mean Squared Displacement}\label{method:realizations}
  %BJC4: reworked this section
  %MRS: should explain why here there are two approaches.
  \textit{Generating Stochastic HDP-AR-HMM Trajectory Realizations}: We used two 
  different methods to stochastically generate trajectories:
  \begin{enumerate}[label={Method \theenumi :}, leftmargin=3.5\parindent]
    \item For each independent MD trajectory, we generated stochastic trajectory
    realizations using the finalized parameters of the associated HDP-AR-HMM. This 
    results in 24 separate ensembles of trajectories for each solute studied. 
    \item Using the clustered parameter sets, we generated a single ensemble
	of trajectories. Realizations are representative of the combined behavior
	of all solute trajectories.
  \end{enumerate}

@@

  %MRS: this doesn't make sense - there aren't 1000 trajectories.  You mean method 1?
  For MD trajectories, we bootstrapped according to Method 2.

@@

  Solute flux, as written, can be expressed in terms of the product of solute 
  diffusivity, $D_i$, within the membrane and a sorption coefficient, $K_i$, 
  that quantifies the ability of the solute to move between the bulk solution
  and and the membrane:
  \begin{equation}
    J_i = \frac{D_i K_i}{l}\Delta C
  \end{equation}
  where $l$ is the membrane thickness. For our purposes, $K_i$ and $l$ are 
  irrelevant since we do not study pore entry effects or flux across finite
  length pores. Therefore, in order to calculate selectivity, with respect
  to quantities we can study with our MD simulations, it is only necessary that
  we compare the ratio of diffusion constants:
  \begin{equation}
    S_{ij} = \frac{D_i}{D_j}
    \label{eqn:selectivity_diffusivity}
  \end{equation}
  It can be shown on time scales where motion is uncorrelated, that the diffusion
  constant is linearly proportional to the slope of the MSD curve:~\cite{einstein_investigations_1956}
  \begin{equation}
    \overline{z^2(\tau)} = 2 n D t
    \label{eqn:msd_D}
  \end{equation}
  where $n$ is the number of spatial dimensions (1 in our case). Since the MSD
  %MRS: tweaked
  %of our models 
  of the realizations of our models 
  enters a linear regime, we can estimate diffusion constants 
  %MRS: maybe can say what the timescale of the linear regime, and how much longer the stochastic realizations are?
  based on the slope of the linear region of the predicted MSD curves.

@@

  \textit{Hydrogen bond lifetimes}: We estimated the lifetime of hydrogen bonds by 
  recording the length of sequential frames where solutes remained hydrogen bonded.
  We still counted sequences where hydrogen bonds were broken for a single frame 
  before reforming. Consistent with our previous work, we reported the 95th 
  percentile of hydrogen bond lifetimes since their distribution is not Gaussian
  and to 
  %MRS: 
  %emphasize 
  accentuate
  longer trapping periods.~\cite{coscia_chemically_2019}

@@

  %We developed a way to 
  \textit{Local density measurements}: We measured the local density of the membrane at
  arbitrary points in the unit cell as follows. We histogrammed the positions in three 
  dimensions. Since our system is in a monoclinic unit cell, we periodically replicated
  the system in the $\pm x$, $y$ and $z$ directions and then chose the bounds on the
  histogram in order to create a rectangular box encompassing the unit cell with
  a 1 nm buffer between the histogram and unit cell boundaries.~\cite{van_der_walt_numpy_2011}
  We then used a regular grid interpolator in order to allow interpolation at arbitrary
  points within the grid.~\cite{virtanen_scipy_2020}

@@

  % BJC4: some kind of brief selectivity analysis here. This easy, I'll get that done 
  % tonight or tomorrow.

>>>>

  Appropriate time series modeling of complex diffusion in soft matter systems on the
  microsecond time scale can provide a path towards inferring transport mechanisms and
  predicting bulk properties characteristic of much longer time scales. In this work 
  we apply the sticky hierarchical Dirichlet process autoregressive hidden Markov model
  (HPD-AR-HMM) to solute center-of-mass trajectories generated from long molecular dynamics 
  simulations in a cross-linked inverted hexagonal phase lyotropic liquid crystal
  (LLC) membrane in order to automatically detect a variety of solute dynamical states
  which can be further explained in terms of solute-membrane interactions.

@@

  %MRS: we have to decide if ``effective'' is the right word. Alternative is ``promising''.
  %BJC5: ``promising'' seems safer
  Overall, this work provides a promising way to simultaneously identify transport 
  mechanisms in nanoporous materials and project complex diffusive behavior on
  long time scales. Our enhanced understanding of the diverse range of solute behavior
  allows us to hypothesize design changes to LLC monomers aimed towards controlling the
  rates of solute passage, thus improving the selective performance of LLC membranes.

@@

  In this work, we are particularly interested in the design of lyotropic liquid 
  crystal (LLC) monomers, a class of amphiphilic molecules whose ordered self-assembled
  phases can be cross-linked into mechanically strong membranes capable of highly 
  selective separations. Inverted hexagonal, H\textsubscript{II}, phase LLC membranes
  are characterized	by hexagonally packed, uniform-sized and straight pores, an ideal
  geometry for high throughput transport. Their pores are lined with the LLC monomer's
  functional groups which can potentially be designed to interact with solutes in a 
  chemically-specific manner. It would be extremely useful to experimentalists if we 
  could connect LLC monomer design with measurable quantities. This goal is complicated
  by the non-Brownian hopping and trapping behavior induced by the membrane's diverse
  topology and inhomogeneous structure resulting in subdiffusive behavior out to 
  multiple microseconds of simulation.~\cite{coscia_understanding_2019,coscia_chemically_2019}

  We have already advanced our understanding of LLC monomer design in our previous work
  by building stochastic time series models based on the chemical intuition gained
  from qualitative MD studies of solute transport in an H\textsubscript{II} phase LLC
  membrane.~\cite{coscia_chemically_2019,coscia_capturing_2020} We identified three 
  classes of trapping behavior which result in subdiffusion out to the microsecond 
  time scale. In our first approach to stochastically model this behavior, we treated 
  the system's dynamics as a sequence of anti-correlated hops between power law distributed periods of 
  entrapment, a framework called subordinated fractional Brownian or L\'evy
  motion.~\cite{thiel_weak_2014,teuerle_modeling_2013} In a second approach, we treated 
  solute motion as a Markov state model~\cite{pande_everything_2010} with state-dependent
  dynamics, where we parameterized the state transition probabilities between each of
  eight discrete states, defined by the observed trapping mechanisms, as well as the 
  solute dynamics within each of these states. We showed how one could use realizations
  of any stochastic model like these in order to predict macroscopic flux and selectivity.

  Although powerful and insightful, our previous work required considerable human effort
  in order to analyze the MD trajectories and to hypothesize stochastic models which
  matched the behavior of solutes. It would be significantly more useful and broadly 
  applicable if we could design an approach which distinguishes and parameterizes varied
  solute dynamics in a data-driven way and leaves a facile way to generate an ensemble
  of stochastic trajectory realizations that can be used to predict macroscopic transport
  properties.

@@

  In their most basic form, hidden Markov models (HMMs) are used to generate
  a series of latent, or hidden, states dependent only on the previous state.~\cite{rabiner_tutorial_1989}
  Propagating an HMM generates sequential but independent observations from 
  the emission distributions associated with each hidden state. Transitions 
  between states are mathematically defined by a matrix of transition probabilities.
  Given a series of observations and a known number of hidden states, the 
  Baum-Welch algorithm, an expectation maximization technique, is the most commonly 
  used way to determine the hidden state sequence and the underlying parameters of
  a standard HMM.~\cite{baum_maximization_1970} It also possible to achieve the same
  results using a Bayesian framework.~\cite{scott_bayesian_2002,jasra_markov_2005} 
  %MRS: not entirely clear what purpose of above sentence its - it sort of implied you will use Bayesian framework?
  % BJC5: Baum welch is most widely used for pure HMM but is an expectation maximization algorithm
  % There are pure Bayesian ways to do it. Tried to clarify above.
  One can place priors on the transition probabilities and parameters of the hidden 
  states' emission distributions in order to iteratively infer their values and to
  estimate the hidden state sequence. HMMs have been used in a broad range of applications
  including speech recognition, weather forecasting and bioinformatics.~\cite{juang_hidden_1984,hughes_non-homogeneous_1999,yoon_hidden_2009}.
  In molecular simulations, HMMs have been used to characterize molecular kinetics
  in order to determine the equilibrium distribution of various observables.~\cite{thayer_hidden_2002,singhal_using_2004,noe_probability_2008,noe_projected_2013}

  By placing a hierarchical Dirichlet process (HDP) prior on the transition 
  probabilities, it is possible to extend HMMs to an unknown number of hidden
  states, thus allowing the model's complexity to be driven by the complexity
  of the data.~\cite{teh_hierarchical_2006} We will refer to this as the HDP-HMM,
  although in literature it is often called the infinite hidden Markov 
  model.~\cite{beal_infinite_2002} The HDP-HMM elegantly overcomes the main
  shortcoming of standard HMMs, the need to choose the number of states 
  \textit{a priori}, making its use more advantageous in a diversity of 
  applications including economic data and single particle trajectory 
  analysis.~\cite{shi_identifying_2016,hines_analyzing_2015}

@@

  %MRS: define what ``this'' is - important claim, make it clear what it is you are claiming.
  To our knowledge, applying the HDP-AR-HMM in order to gain kinetic and
  mechanistic insight from MD simulation trajectories has not been done
  before.
  % this is a novel application and analysis of the sticky HDP-AR-HMM.
  Typical applications of the HDP-AR-HMM center around determining the 
  number of states, the state sequence and occasional analysis of the emission 
  distributions of a time series. The addition of temporal correlation to 
  the inference procedure has primarily been used to improve state sequence 
  identification.~\cite{calderon_inferring_2015,hamada_modeling_2016}
  Not only do we use the HDP-AR-HMM to segment MD time series into distinct
  dynamical modes, but we cluster and analyze the temporally-correlated 
  parameters of the model in order to learn physical mechanisms and describe
  solute kinetics. Further, we use the model parameters to generate new 
  time series trajectories which enables projections of MD simulations on 
  much longer time scales.

@@

  We ran all MD simulations and energy minimizations using GROMACS 2018.~\cite{bekker_gromacs:_1993,berendsen_gromacs:_1995,van_der_spoel_gromacs:_2005,hess_gromacs_2008}  
  We implemented the sticky HDP-AR-HMM in Python 
  (\texttt{https://github.com/bencoscia/hdphmm}) which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_nonparametric_2009} Since we do not incorporate
  any novel methodology to the HDP-AR-HMM inference procedure, we refer the 
  interested reader to an excellent and in-depth description of the formulation of
  the HDP-AR-HMM and related models.~\cite{fox_bayesian_2010} 
  %MRS: possibly should give the permanent reference to a shirtsgroup fork?  We can discuss which is likely to be more permenant  
  All other post-simulation trajectory analysis tools are available online at
  \texttt{https://github.com/shirtsgroup/LLC\_Membranes}.

@@

  Each solute system was simulated for 5 $\mu$s of molecular dynamics simulation. We
  used a time step of 2 fs at a pressure of 1 bar and temperature of 300K controlled
  by the semi-isotropic Parrinello-Rahman 
  %MRS: just wondering, was this semiisotropic, or isotropic?  If isotropic, do we know how different the diagonals of the pressure tensor were (since the system is not totally isotropic)
  %BJC5: I use semi-isotropic (xy together and z separate)
  barostat~\cite{parrinello_polymorphic_1981} and velocity rescale 
  thermostat~\cite{bussi_canonical_2007} respectively. We recorded configurations every 0.5 ns.

@@

  We use the sticky HDP-AR-HMM in order to assign the time-correlated positional 
  fluctuations of solute center-of-mass MD trajectories to discrete dynamical modes while 
  simultaneously inferring the VAR parameters of those modes. We use the sticky model
  in order to emphasize self-transitions since we expect solutes to stay trapped in
  states for appreciable amounts of time. 
%  We analyze solute motion in terms of their
%  cylindrical coordinates, $(r, z)$, with the $z$ axis oriented along the membrane 
%  pores and the radial dimension, $r$, measured from the closest pore center.

  Although the discussion will be presented in terms of 2D cylindrical coordinates, we
  applied the HDP-AR-HMM algorithm to 3D solute center-of-mass coordinate trajectories 
  with the radial dimension represented by the solute's $x$ and $y$ coordinates 
  transformed relative to the nearest pore center, and the $z$ axis oriented along 
  the membrane pore. By working in Cartesian coordinates, we avoid mathematical 
  complexity introduced by change of coordinates into cylindrical coordinates while
  estimating the state sequence.

@@

  We clustered based on the eigenvalues of $A$ and $\Sigma$, the distance from the 
  pore center and the self-transition probabilities, $T_{ii}$, of each state. $A$ 
  and $\Sigma$ capture the correlation and size of fluctuations. $T_{ii}$ is directly
  related to the average time spent in each state. The distance from the closest pore 
  center is likely meaningful since the membrane is radially inhomogeneous from each 
  pore, transitioning from hydrophilic pores to hydrophobic tails. However, the 
  channels are isotropic along the pore axis so we do not cluster in means in the 
  $z$ dimension. We normalized each clustering dimension by its variance and shifted
  the mean to zero.

  To take advantage of the system's cylindrical symmetry, we converted the parameters that
  are directly related to solute position into cylindrical coordinates before clustering. 
  We replaced the $x$ and $y$ eigenvalues of $\Sigma$ and $A$ with $\lambda_x$ + 
  $\lambda_y$ and $\lambda_x^2$ + $\lambda_y^2$ respectively, which are invariant with 
  rotation around the $z$ axis. We clustered on squared means in $r$, 
  $\mu_r^2 = \mu_x^2 + \mu_y^2$, rather than $\mu_x$ and $\mu_y$ individually, with the
  square $\mu_r$ converting it into the same units as $A$ and $\Sigma$.

  To cluster on $T_{ii}$, we first cast it in terms of the expected dwell times. The expected
  value of the number of sequential self-transitions is simply:
  \begin{equation}  
   E[dwell~time] = \frac{1}{1 - T_{ii}}
   \label{eqn:dwell_times}
  \end{equation}  
  This relationship implies that dwell times approach infinity as $T_{ii}$ approaches
  1. Therefore we found the most success with agglomerative clustering by linearizing this
  relationship and clustering on $-\log(1 - T_{ii})$.

  There are alternative ways to cluster the data dependent on one's goals. Our goal is
  to derive the simplest set of parameter which adequately describe solute motion. 
  However, one may want a more detailed understand of specific dimensions of the 
  clustering dataset. In this case one can cluster on one or more dimensions independently
  and then enumerate the possible clusters. This would ensure very little overlap of
  clusters but may result in a large number of them. Once one has achieved a better 
  understanding of solute-membrane interactions, perhaps by using the HDP-AR-HMM, it may be
  beneficial to incorporate other physically motivated variables into the clustering
  procedure. 
  %For example, one can gain a thorough understanding of the dynamics of
  %different kinds of trapping behavior related to hydrogen bonding and sodium ion association.
  %MRS: last sentence is a little weird. Audience may be left without knowing how this could be achieved.

@@

  \textit{Obtaining Parameters of Clustered States}: We remapped and fixed the
  state sequence based on the cluster assignments and 
  parameterized the clusters using the inference component of the HDP-AR-HMM algorithm.
  First, we modified the ($x$, $y$, $z$) solute trajectories so that they had a 
  mean of zero, leaving only the fluctuations. Next, we took advantage of the
  rotational invariance of the system by rotating the coordinates about the pore 
  axis such that the mean vector of each separate segment was aligned with the 
  $x$ axis. Next, we subtracted the means from each separate segment in the $x$ 
  and $z$ dimensions, with the mean in $y$ being zeroed by the rotation. We used
  the inference component of the HDP-AR-HMM algorithm on this modified trajectory to 
  infer the clustered state parameters by fixing the clustered state sequence.

@@

  \subsection{Predicting Mean Squared Displacement}\label{method:realizations}
  %BJC4: reworked this section
  %MRS: should explain why here there are two approaches.
  \textit{Generating Stochastic HDP-AR-HMM Trajectory Realizations}: We used two 
  different methods to stochastically generate trajectories because generating
  trajectories from clustered parameter sets can give results that are qualitatively 
  and quantitatively different from those generated by unclustered parameter sets.
  In our discussion we evaluate both methods:
  \begin{enumerate}[label={Method \theenumi :}, leftmargin=3.5\parindent]
    \item For each independent MD trajectory, we generated stochastic trajectory
    realizations using the finalized parameters of the associated HDP-AR-HMM. This 
    results in 24 separate ensembles of trajectories for each solute studied. 
    \item Using the clustered parameter sets, we generated a single ensemble
	of trajectories. Realizations are representative of the combined behavior
	of all solute trajectories.
  \end{enumerate}

@@

  For the MD MSDs, we bootstrapped according to Method 1, sampling with 
  replacement from the ensemble of 24 solute trajectories.

@@

  Solute flux, as written, can be expressed in terms of the product of solute 
  diffusivity, $D_i$, within the membrane and a sorption coefficient, $K_i$, 
  that quantifies the ability of the solute to move between the bulk solution
  and and the membrane:
  \begin{equation}
    J_i = \frac{D_i K_i}{l}\Delta C
  \end{equation}
  where $l$ is the membrane thickness. For our purposes, $K_i$ and $l$ are 
  irrelevant since we do not study pore entry effects or flux across finite
  length pores. Therefore, in order to calculate selectivity, with respect
  to quantities we can study with our MD simulations, it is only necessary that
  we compare the ratio of diffusion constants:
  \begin{equation}
    S_{ij} = \frac{D_i}{D_j}
    \label{eqn:selectivity_diffusivity}
  \end{equation}
  It can be shown on time scales where motion is uncorrelated, that the diffusion
  constant is linearly proportional to the slope of the MSD curve:~\cite{einstein_investigations_1956}
  \begin{equation}
    \overline{z^2(\tau)} = 2 n D t
    \label{eqn:msd_D}
  \end{equation}
  where $n$ is the number of spatial dimensions (1 in our case). Since the MSD
  of the realizations of our models enters a linear regime, we can estimate 
  diffusion constants 
  % MRS: maybe can say what the timescale of the linear regime, and how much 
  % longer the stochastic realizations are?
  based on the slope of the linear region of the predicted MSD curves.

@@

  \textit{Hydrogen bond lifetimes}: We estimated the lifetime of hydrogen bonds by 
  recording the length of sequential frames where solutes remained hydrogen bonded.
  We still counted sequences where hydrogen bonds were broken for a single frame 
  before reforming. Consistent with our previous work, we reported the 95th 
  percentile of hydrogen bond lifetimes since their distribution is not Gaussian
  and to accentuate longer trapping periods.~\cite{coscia_chemically_2019}

@@

  \textit{Local density measurements}: We measured the local density of the membrane at
  arbitrary points in the unit cell as follows. We histogrammed the positions in three 
  dimensions. Since our system is in a monoclinic unit cell, we periodically replicated
  the system in the $\pm x$, $y$ and $z$ directions and then chose the bounds on the
  histogram in order to create a rectangular box encompassing the unit cell with
  a 1 nm buffer between the histogram and unit cell boundaries.~\cite{van_der_walt_numpy_2011}
  We then used a regular grid interpolator in order to allow interpolation at arbitrary
  points within the grid.~\cite{virtanen_scipy_2020}

@@

  % BJC4: some kind of brief selectivity analysis here.
